# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r, echo=TRUE}
library(ggplot2)
library(lubridate)
activity <- read.csv("activity.csv")
summary(activity)
activity$date = as.Date(activity$date)
unique(activity$date)
```

## What is mean total number of steps taken per day?

```{r, echo=TRUE}
tot_steps <- tapply(activity$steps, activity$date, sum)
qplot(tot_steps, main = "Total steps taken per day",
       xlab = "Total steps", ylab = "Count", margins = T)

## calculate the mean and median total number of steps taken per day
act_mean <- mean(tapply(activity$steps, activity$date, sum), 
                 na.rm = TRUE)
act_median <- median(tapply(activity$steps, activity$date, sum), 
                     na.rm = TRUE)
act_mean
act_median
## mean is 10766.19, median is 10765
```

## What is the average daily activity pattern?

```{r, echo=TRUE}
intervals <- unique(activity$interval)
avgsteps <- tapply(activity$steps, activity$interval, mean, na.rm = T)
interval_data <- data.frame(intervals, avgsteps)
with(interval_data, plot(intervals, avgsteps, 
                         main = "Average daily steps per interval", 
                         ylab = "average steps", type = "l"))
max_index <- as.numeric(which.max(interval_data$avgsteps))
interval_data[max_index,]
## the 835 interval has the most with 206 steps
```

## Imputing missing values

First calulating the total number of NAs:
```{r, echo=TRUE}
table(is.na(activity$steps))
```

For missing values, replacing them with the mean at that same interval, as I am assuming that activities follow a daily pattern.
```{r, echo=TRUE}
steps_per <- aggregate(activity$steps, 
                       by = list(interval = activity$interval), 
                       mean, na.rm = T)
# convert to integers for plotting
steps_per$interval <- as.integer(levels(steps_per$interval)
                                [steps_per$interval])
colnames(steps_per) <- c("interval", "steps")

means_replace <- function(activity, defaults) {
    na_indices <- which(is.na(activity$steps))
    defaults <- steps_per
    na_repl <- unlist(lapply(na_indices, 
                                     FUN = function(ind) {
        interval = activity[ind, ]$interval
        defaults[defaults$interval == interval, ]$steps
    }))
    new_steps <- activity$steps
    new_steps[na_indices] <- na_repl
    new_steps
}
activity_imp <- data.frame(steps = means_replace(activity, avgsteps), 
                           date = activity$date, 
                           interval = activity$interval)

summary(activity_imp)
```

Plotting new dataset with inserted steps:
```{r, echo=TRUE}
intervals <- unique(activity_imp$interval)
avgsteps <- tapply(activity_imp$steps, activity_imp$interval, 
                   mean, na.rm = T)
interval_data_imp <- data.frame(intervals, avgsteps)
with(interval_data_imp, plot(intervals, avgsteps, 
                         main = "Average daily steps per interval", 
                         ylab = "average steps", type = "l")) 
```

Calculating the new mean and median
```{r, echo=TRUE}
imp_mean <- mean(tapply(activity_imp$steps, activity_imp$date, sum), 
                 na.rm = TRUE)
imp_median <- median(tapply(activity_imp$steps, activity_imp$date, sum), 
                     na.rm = TRUE)
imp_mean
imp_median
```
